### 2016-04-26 text/Chapter05.tex
Edits (Chapter 5)

The \emph{flat coeffect calculus} introduced in the previous chapter uniformly captures a number
of context-aware systems outlined in Chapter~\ref{ch:applications}. The coeffect calculus
can be seen as a \emph{language framework} that simplifies the construction of concrete
\emph{domain-specific} coeffect lnguages. In the previous chapter, we discussed how it provides a
type system that tracks the required context. In this chapter, we show that the langauge framework
also provides a way for defining the semantics of concrete domain-specific coeffect languages,
guides their implementation and simplifies safety proofs.

@@

We use comonads in a syntactic way, following the example of Wadler and Thiemann \cite{monads-effects-marriage}
and Haskell's use of monads. The translation is the same for all coeffect languages, but the safety
depends on the concrete coeffect-specific comonadically-inspired primitives. We prove the soundness
of two concrete coeffect calculi (dataflow and implicit parameters). We note that the proof
crucially relies on a relationship between coeffect annotations (provided by the type system) and
the comonadically-inspired primitives (defining the semantics), which makes it easy to extend it to
other concrete context-aware langauges.

@@

To provide an intuition, consider two context-aware programs. The first calls a function that
adds two implicit parameters in a context where one of them is defined. The second calculates
the difference between the current and the previous value in a dataflow computation.
For comparison, we show the code written in a coeffect dataflow language (on the left) and
using standard ML-like libraries (on the right):
%
\begin{equation*}
\begin{array}{lcl}
\begin{array}{l}
\kvd{let}~\ident{add}=\kvd{fun}~\ident{x}\rightarrow\\[-0.2em]
 \quad \ident{?one} + \ident{?two}~\kvd{in}\\[-0.2em]
\kvd{let}~\ident{?one} = 10~\kvd{in}\\[-0.2em]
\ident{add}~0
\end{array}
&\quad&
\begin{array}{l}
\kvd{let}~\ident{add}=\kvd{fun}~\ident{x}~\ident{params}\rightarrow\\[-0.2em]
 \quad \ident{lookup}~\str{one}~\ident{params}~+ \\[-0.2em]
 \quad\quad \ident{lookup}~\str{two}~\ident{params}~\kvd{in}\\[-0.2em]
\ident{add}~0~(\ident{cons}~\str{one}~10~\ident{params})
\end{array}
\\[3.5em]
\begin{array}{l}
\kvd{let}~\ident{diff}=\kvd{fun}~\ident{x}\rightarrow\\[-0.2em]
 \quad \ident{x}-\kvd{prev}~\ident{x}
\end{array}
&&
\begin{array}{l}
\kvd{let}~\ident{diff}=\kvd{fun}~\ident{x}\rightarrow\\[-0.2em]
 \quad \ident{List.head}~\ident{x}-\ident{List.head}~(\ident{List.tail}~\ident{x})
\end{array}
\end{array}
\end{equation*}
%
The \ident{add} function (on the left) has a type $\ident{int}\xrightarrow{\cclrd{ \{ \ident{?one},\ident{?two} \} }}\ident{int}$.
We call it in a context containing $\cclrd{\ident{?one}}$ and so the coeffect of the program is
$\cclrd{ \{ \ident{?two} \} }$. The safety property for implicit parameters
(Theorem~\ref{thm:semantics-sound-impl}) guarantees that, when executed in a context that
provides a value for the implicit parameter $\cclrd{\ident{?one}}$, the program reduces to
a value of the correct type (or never terminates).

@@

It is also interesting to examine some data types that do \emph{not} form a comonad. As already
mentioned, list $\ident{List}~\alpha = 1 + (\alpha \times \ident{List}~\alpha)$ is not a comonad,
because the \ident{counit} operation is not defined for the value $\kvd{inl}~()$. The \ident{Maybe}
data type defined as $1 + \alpha$ is not a comonad for the same reason. However, if we consider
flat coeffect calculus for liveness, it appears natural to model computations as functions
$\ident{Maybe}~\tau_1 \rightarrow \tau_2$. To use such a model, we need to generalise comonads
to \emph{indexed comonads}.

@@

\noindent
The $\ident{merge}_{\cclrd{r},\cclrd{s}}$ operation is the most interesting one. Given two comonadic
values with additional contexts specified by $\cclrd{r}$ and $\cclrd{s}$, it combines them into a
single value with additional context $\cclrd{r}\,\czip\,\cclrd{s}$. The $\czip$ operation often represents
\emph{greatest lower bound}\footnote{The $\czip$ and $\cpar$ operations are the greatest and least upper
bounds in the liveness and data-flow examples, but not for implicit parameters. However, they remain useful
as an informal analogy.}. We look at examples of this operation in the next section.

The $\ident{split}_{\cclrd{r},\cclrd{s}}$ operation splits a single comonadic value (containing a tuple)
into two separate values. Note that this does not simply duplicate the value, because the additional
context is also split. To obtain coeffects $\cclrd{r}$ and $\cclrd{s}$, the input needs to provide
\emph{at least} $\cclrd{r}$ and $\cclrd{s}$, so the tags are combined using the $\cpar$, which is often
the \emph{least upper-bound}\footnotemark[1].

\paragraph{Semantics of sub-coeffecting.}
Although we do not include sub-coeffecting in the core flat coeffect calculus, it is an interesting
extension to consider. Semantically, sub-coeffecting drops some of the available contextual
capabilities (drops some of the implicit parameters or some of the past values). This can be
modelled by adding a (family of) lifting operation(s):
%
\begin{itemize}
 \item $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ is a family of mappings
   $\ctyp{\cclrd{r'}}{\alpha} \rightarrow \ctyp{\cclrd{r}}{\alpha}$~ for all $\cclrd{r'}, \cclrd{r}$ such that $\cclrd{r}\,\cleq\,\cclrd{r'}$
\end{itemize}
%
Although we do not demand this as a general law, in all our systems, it is the case that
$\cclrd{r} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$ and $\cclrd{s} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$.
This special case allows a simpler definition of \emph{indexed flat comonad} by expressing the
\ident{split} operation in terms of lifting (sub-coeffecting) as follows:
%
\begin{equation*}
\begin{array}{rcl}
\ident{map}_{\cclrd{r}}~f &\narrow{=}& \ident{cobind}_{\cclrd{r}, \cclrd{r}}~(f\circ\ident{counit}_{\cunit}) \\
\ident{split}_{\cclrd{r}, \cclrd{s}}~c &\narrow{=}&
  ( \ident{map}_{\cclrd{r}}~\ident{fst}~(\ident{lift}_{\cclrd{r}\cpar\cclrd{s}, \cclrd{r}}~c),
    \ident{map}_{\cclrd{s}}~\ident{snd}~(\ident{lift}_{\cclrd{r}\cpar\cclrd{s}, \cclrd{s}}~c) )
\end{array}
\end{equation*}
%
The $\ident{map}_{\cclrd{r}}$ operation is the mapping on arrows that corresponds to the object
mapping $\ctyp{\cclrd{r}}{}$. The definition is dual to the standard definition of \ident{map}
for monads in terms of \ident{bind} and \ident{unit}. The functions \ident{fst} and \ident{snd}
are first and second projections from a two-element pair. To define the
$\ident{split}_{\cclrd{r}, \cclrd{s}}$ operation, we use the argument $c$ twice, use lifting
to throw away additional parts of the context and then transform the values in the
context.

This alternative is valid for our examples, but we do not use it for three reasons. First, it
requires making sub-coeffecting a part of the core definition. Second, this would be the only
place where our semantics uses a variable \emph{twice} (in this case $c$). Thus using an explicit
\ident{split} means that the structure required by our semantics does not need to provide variable
duplication and our model could be embedded in linear or affine category. Finally, explicit
\ident{split} is similar to the definition that is needed for structural coeffects in
Chapter~\ref{ch:structural} and it makes the connection between the two easier to see.

\paragraph{Examples.}
All the examples of \emph{indexed comonads} discussed in Section~\ref{sec:semantics-flat-idx} can
be extended into \emph{flat indexed comonads}. Note that this is not true \emph{in general},
because each example requires us to define additional operations, specific for the example.

\begin{example}[Monoidal comonads]
Just like indexed comonads generalise co\-monads, the additional structure of
flat indexed comonads generalises symmetric semimonoidal comonads of Uustalu
and Vene \cite{comonads-notions}. The flat coeffect algebra is defined as $(\{1\}, \ast, \ast, \ast, 1, 1, =)$
where $1\ast1=1$ and $1=1$. The additional operation $\ident{merge}_{1,1}$ is provided by the
monoidal operation called \ident{m} by Uustalu and Vene. The $\ident{split}_{1,1}$ operation
is defined by duplication.
\end{example}

\begin{example}[Indexed Maybe]
\label{thm:semantics-indexed-opt}

Flat coeffect algebra for liveness defines $\cpar$ and $\czip$ as $\sqcup$ and $\sqcap$, respectively
and specifies that $\ident{D} \sqsubseteq \ident{L}$. Recall also that the object mapping is defined
as $\ctyp{\ident{L}}{\alpha} = \alpha$ and $\ctyp{\ident{D}}{\alpha} = 1$. The additional operations
of a flat indexed comonad are defined as follows:
%
\begin{equation*}
\begin{array}{rcl}
\ident{merge}_{\ident{L}, \ident{L}}~(a, b) &\narrow{=}& (a, b)\\
\ident{merge}_{\ident{L}, \ident{D}}~(a, ()) &\narrow{=}& ()\\
\ident{merge}_{\ident{D}, \ident{L}}~((), b) &\narrow{=}& ()\\
\ident{merge}_{\ident{D}, \ident{D}}~((), ()) &\narrow{=}& ()\\
\end{array}
\qquad
\begin{array}{rcl}
\ident{split}_{\ident{L}, \ident{L}}~(a, b) &\narrow{=}& (a, b)\\
\ident{split}_{\ident{L}, \ident{D}}~(a, b) &\narrow{=}& (a, ())\\
\ident{split}_{\ident{D}, \ident{L}}~(a, b) &\narrow{=}& ((), b)\\
\ident{split}_{\ident{D}, \ident{D}}~() &\narrow{=}& ((), ()))\\
\end{array}
\end{equation*}
\end{example}

@@

For implicit parameters, both $\czip$ and $\cpar$ are the $\cup$ operation and the relation
$\cleq$ is formed by the subset relation $\subseteq$. Recall that the data type $\ctyp{\cclrd{r}}{\alpha}$
is $\alpha \times (\cclrd{r} \rightarrow \ident{num})$ where $\ident{num}$ is the type of implicit parameter values.
The additional operations are defined as:
%
\begin{equation*}
\begin{array}{rcl}
\ident{split}_{\cclrd{r}, \cclrd{s}}~((a,b), g) &\narrow{=}& ((a, \restr{g}{\cclrd{r}}), (b, \restr{g}{\cclrd{s}}))\\
\ident{merge}_{\cclrd{r}, \cclrd{s}}~((a, f), (b, g)) &\narrow{=}& ((a, b), f \uplus g)\\
\end{array}
\quad
\begin{array}{l}
\textnormal{where}~f \uplus g = \\[-0.25em]
\quad\restr{f}{\, \textit{dom}(f) \setminus \textit{dom}(g)} \cup g
\end{array}
\end{equation*}
\end{example}

\noindent
The \ident{split} operation splits the tuple and restricts the function (representing available
implicit parameters) to the required sub-sets. The \ident{merge} operation is more
interesting. It uses the $\uplus$ operation that we defined when introducing implicit parameters
in Section~\ref{sec:applications-flat-impl}. It merges the values, preferring the definitions from
the right-hand side (call site) over left-hand side (declaration site). Thus the operation is not
symmetric.

@@

The $\ident{cobind}_{m,n}$ operation requires $m + n$ elements in order to generate $n$ past results
of the $f$ function, which itself requires $m$ past values. When combining two lists,
$\ident{merge}_{m,n}$ behaves as \emph{zip} and produces a list that has the length of the shorter
argument. When splitting a list, $\ident{split}_{m, n}$ needs the maximum of the required lengths.
Finally, the lifting operation just drops some number of elements from a list.

@@

\begin{figure*}[t]
The semantics is defined over a typing derivation:
%
\begin{equation*}
\hspace{-3em}
\begin{array}{ll}
\hspace{5.2em}\semdef
  {\coctx{\Gamma}{\cunit} \vdash x_i : \tau_i }
  {\pi_i \circ \ident{counit}_{\cunit}}
& (\emph{var})
\\[1em]
\hspace{4.9em}\semdef
  {\coctx{\Gamma}{\czero} \vdash n : \ident{num} }
  {\ident{const}~n}
& (\emph{num})
\\[1.5em]
\hspace{2.2em}\semdeff
  {\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
  {f}
  {f \circ \ident{curry}~\ident{merge}_{\cclrd{r}, \cclrd{s}} }
& (\emph{abs})
\\[2em]
\semdefff
  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}
  {\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
  {\begin{array}{l}\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2\\[-0.3em]~\end{array}}
  {f}
  {g}
  {\hspace{-0.5em}\begin{array}{l}
  \ident{app}~\circ~f\hspace{-0.15em}\times\hspace{-0.15em}(\ident{cobind}_{\cclrd{s}, \cclrd{t}}~g)~\circ~\ident{split}_{\cclrd{r}, \cclrd{s} \,\cseq\, \cclrd{t}} \\[-0.3em]
  ~~\circ~\ident{map}_{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})}~\ident{dup}\\[-0.3em]
  \end{array}\hspace{-0.5em}}
& (\emph{app})
\end{array}
\end{equation*}

@@

In Section~\ref{sec:applications-flat}, we defined the semantics of concrete (flat) context-dependent
computations including implicit parameters, liveness and data-flow. Using the \emph{flat indexed
comonad} structure, we can now define a single uniform semantics that is capable of capturing all
our examples, as well as other computations that can be modelled by the structure.

As discussed in Section~\ref{sec:flat-unique}, different typing derivations of coeffect programs
may have different meaning (\eg~when working with implicit parameters) and so the semantics is
defined over a \emph{typing derivation} rather than over an \emph{term}. To assign a
semantics to a term, we choose its unique coeffect-specific typing derivation as defined in
Section~\ref{sec:flat-unique}.

\paragraph{Contexts and types.}
The modelling of contexts and functions generalizes the earlier concrete examples. We use the
family of mappings $\ctyp{\cclrd{r}}{}$ as an (indexed) data-type that wraps the product of
free variables of the context and the arguments of functions:
%
\begin{equation*}
\begin{array}{rcl}
\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \cclrd{r} } \vdash e : \tau}
  &:& \ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau\\
\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2} &=& \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2
\end{array}
\end{equation*}

@@

The semantics of variable access and abstraction are the same as in the semantics of Uustalu and
Vene \cite{comonads-notions}, modulo the indexing. The semantics of variable access (\emph{var}) uses
$\ident{counit}_{\cunit}$ to extract a product of free variables and projection
$\pi_i$ to obtain the variable value. Abstraction (\emph{abs}) is interpreted as a curried
function that takes the declaration-site context and a function argument, merges them using
$\ident{merge}_{\cclrd{r}, \cclrd{s}}$ and passes the result to the semantics of the body $f$.
Assuming the context $\Gamma$ contains variables of types $\sigma_1, \ldots, \sigma_n$, this gives
us a value $\ctyp{\cclrd{r}\czip\cclrd{s}}((\sigma_1 \times \ldots \times \sigma_n) \times \tau_1)$.
Assuming that $n$-element tuples are associated to the left, the wrapped context is equivalent to
$\sigma_1 \times \ldots \times \sigma_n \times \tau_1$, which can then be passed to the body of the
function.

@@

\paragraph{Properties.}
We the categorical semantics  in Section~\ref{sec:semantics-translation} to define a translation
that embeds context-dependent computations in a functional programming langauge, similarly to how
monads and the do notation provide a way of embedding effectful computations in Haskell.

@@

Informally, the following states that if we see the semantics as a translation, the resulting code
is well-typed. We revisit the property in Lemma~\ref{thm:semantics-welltyped} once we define
the target langauge and its typing.

\begin{lemma}[Correspondence]
\label{thm:flat-correspondence}
In the semantics defined in Figure~\ref{fig:semantics-flat}, the context annotations $\cclrd{r}$ of
typing judgements $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ and function types
$\tau_1 \xrightarrow{\cclrd{r}} \tau_2$  on the left-hand side correspond to the indices of mappings
$\ctyp{\cclrd{r}}{}$ in the corresponding semantic function on the right-hand side.
\end{lemma}
\begin{proof}
By analysis of the semantic rules in Figure~\ref{fig:semantics-flat}.
\end{proof}

@@

\caption{Common syntax and reduction rules of the target langauge}
\label{fig:semantics-target}
\end{figure}

@@

\section{Embedding coeffect languages}
\label{sec:semantics-translation}

@@

\subsection{Functional target langauge}
\label{sec:semantics-translation-target}

>>>>

The \emph{flat coeffect calculus} introduced in the previous chapter uniformly captures a number
of context-aware systems outlined in Chapter~\ref{ch:applications}. The coeffect calculus
can be seen as a \emph{language framework} that simplifies the construction of concrete
\emph{domain-specific} coeffect lnguages. In the previous chapter, we discussed how it provides a
type system that tracks the required context. In this chapter, we show that the language framework
also provides a way for defining the semantics of concrete domain-specific coeffect languages,
guides their implementation and simplifies safety proofs.

@@

We use comonads in a syntactic way, following the example of Wadler and Thiemann \cite{monads-effects-marriage}
and Haskell's use of monads. The translation is the same for all coeffect languages, but the safety
depends on the concrete coeffect-specific comonadically-inspired primitives. We prove the soundness
of two concrete coeffect calculi (dataflow and implicit parameters). We note that the proof
crucially relies on a relationship between coeffect annotations (provided by the type system) and
the comonadically-inspired primitives (defining the semantics), which makes it easy to extend it to
other concrete context-aware languages.

@@

To provide an intuition, consider two context-aware programs. The first calls a function that
adds two implicit parameters in a context where one of them is defined. The second calculates
the difference between the current and the previous value in a dataflow computation.
For comparison, we show the code written in a coeffect dataflow language (on the left) and
using standard ML-like libraries (on the right):
%
\begin{equation*}
\begin{array}{lcl}
\begin{array}{l}
\kvd{let}~\ident{add}=\kvd{fun}~\ident{x}`\rightarrow`\\[-0.2em]
 \quad \ident{?one} + \ident{?two}~\kvd{in}\\[-0.2em]
\kvd{let}~\ident{?one} = 10~\kvd{in}\\[-0.2em]
\ident{add}~0
\end{array}
&\quad&
\begin{array}{l}
\kvd{let}~\ident{add}=\kvd{fun}~\ident{x}~\ident{params}\rightarrow\\[-0.2em]
 \quad \ident{lookup}~\str{one}~\ident{params}~+ \\[-0.2em]
 \quad\quad \ident{lookup}~\str{two}~\ident{params}~\kvd{in}\\[-0.2em]
\ident{add}~0~(\ident{cons}~\str{one}~10~\ident{params})
\end{array}
\\[3.5em]
\begin{array}{l}
\kvd{let}~\ident{diff}=\kvd{fun}~\ident{x}\rightarrow\\[-0.2em]
 \quad \ident{x}-\kvd{prev}~\ident{x}
\end{array}
&&
\begin{array}{l}
\kvd{let}~\ident{diff}=\kvd{fun}~\ident{x}\rightarrow\\[-0.2em]
 \quad \ident{List.head}~\ident{x}-\ident{List.head}~(\ident{List.tail}~\ident{x})
\end{array}
\end{array}
\end{equation*}
%
The \ident{add} function (on the left) has a type $\ident{int}\xrightarrow{\cclrd{ \{ \ident{?one},\ident{?two} \} }}\ident{int}$.
We call it in a context containing $\cclrd{\ident{?one}}$ and so the coeffect of the program is
$\cclrd{ \{ \ident{?two} \} }$. The safety property for implicit parameters
(Theorem~\ref{thm:semantics-sound-impl}) guarantees that, when executed in a context that
provides a value for the implicit parameter $\cclrd{\ident{?one}}$, the program reduces to
a value of the correct type (or never terminates).

@@

It is also interesting to examine some data types that do \emph{not} form a comonad. As already
mentioned, list $\ident{List}~\alpha = 1 + (\alpha \times \ident{List}~\alpha)$ is not a comonad,
because the \ident{counit} operation is not defined for the value $\kvd{inl}~()$. The \ident{Maybe}
data type defined as $1 + \alpha$ is not a comonad for the same reason. However, if we consider
flat coeffect calculus for liveness, it appears natural to model computations as functions
$\ident{Maybe}~\tau_1 \rightarrow \tau_2$. To use such a model, we need to generalize comonads
to \emph{indexed comonads}.

@@

\noindent
The $\ident{merge}_{\cclrd{r},\cclrd{s}}$ operation is the most interesting one. Given two comonadic
values with additional contexts specified by $\cclrd{r}$ and $\cclrd{s}$, it combines them into a
single value with additional context $\cclrd{r}\,\czip\,\cclrd{s}$. The $\czip$ operation often represents
\emph{greatest lower bound}. We look at examples of this operation in the next section.

The $\ident{split}_{\cclrd{r},\cclrd{s}}$ operation splits a single comonadic value (containing a tuple)
into two separate values. Note that this does not simply duplicate the value, because the additional
context is also split. To obtain coeffects $\cclrd{r}$ and $\cclrd{s}$, the input needs to provide
\emph{at least} $\cclrd{r}$ and $\cclrd{s}$, so the tags are combined using the $\cpar$, which is often
the \emph{least upper-bound}\footnote{The $\czip$ and $\cpar$ operations are the greatest and least upper
bounds in the liveness and data-flow examples, but not for implicit parameters. However, they remain useful
as an informal analogy.}.

\paragraph{Semantics of sub-coeffecting.}
Although we do not include sub-coeffecting in the core flat coeffect calculus, it is an interesting
extension to consider. Semantically, sub-coeffecting drops some of the available contextual
capabilities (drops some of the implicit parameters or some of the past values). This can be
modelled by adding a (family of) lifting operation(s):
%
\begin{itemize}
 \item $\ident{lift}_{\cclrd{r'},\cclrd{r}}$ is a family of mappings
   $\ctyp{\cclrd{r'}}{\alpha} \rightarrow \ctyp{\cclrd{r}}{\alpha}$~ for all $\cclrd{r'}, \cclrd{r}$ such that $\cclrd{r}\,\cleq\,\cclrd{r'}$
\end{itemize}
%
The axioms of flat coeffect algebra do not, in general, require that
$\cclrd{r} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$ and $\cclrd{s} \;\cleq\; \cclrd{r}\,\cpar\,\cclrd{s}$,
but the property holds for the three sample coeffect systems we consier. For systems with
the above property, the \ident{split} operation can be expressed in terms of lifting
(sub-coeffecting) as follows:
%
\begin{equation*}
\begin{array}{rcl}
\ident{map}_{\cclrd{r}}~f &\narrow{=}& \ident{cobind}_{\cclrd{r}, \cclrd{r}}~(f\circ\ident{counit}_{\cunit}) \\
\ident{split}_{\cclrd{r}, \cclrd{s}}~c &\narrow{=}&
  ( \ident{map}_{\cclrd{r}}~\ident{fst}~(\ident{lift}_{\cclrd{r}\cpar\cclrd{s}, \cclrd{r}}~c),~
    \ident{map}_{\cclrd{s}}~\ident{snd}~(\ident{lift}_{\cclrd{r}\cpar\cclrd{s}, \cclrd{s}}~c) )
\end{array}
\end{equation*}
%
The $\ident{map}_{\cclrd{r}}$ operation is the mapping on arrows that corresponds to the object
mapping $\ctyp{\cclrd{r}}{}$. The definition is dual to the standard definition of \ident{map}
for monads in terms of \ident{bind} and \ident{unit}. The functions \ident{fst} and \ident{snd}
are first and second projections from a two-element pair. To define the
$\ident{split}_{\cclrd{r}, \cclrd{s}}$ operation, we use the argument $c$ twice, use lifting
to throw away additional parts of the context and then transform the values in the
context.

This alternative definition is valid for our examples, but we do not use it for three reasons. First,
it requires making sub-coeffecting a part of the core definition. Second, this would be the only
place where our semantics uses a variable \emph{twice} (in this case $c$). Note therefore that
our use of an explicit \ident{split} means that the structure required by our semantics does not
need to provide variable duplication and our model could be embedded in linear or affine category.
Finally, explicit \ident{split} is similar to the definition that is needed for structural coeffects
in Chapter~\ref{ch:structural} and it makes the connection between the two easier to see.

\paragraph{Examples.}
All the examples of \emph{indexed comonads} discussed in Section~\ref{sec:semantics-flat-idx} can
be extended into \emph{flat indexed comonads}. Note however that this cannot be done mechanically,
because each example requires us to define additional operations, specific for the example.

\begin{example}[Monoidal comonads]
Just like indexed comonads generalize co\-monads, the additional structure of
flat indexed comonads generalizes the symmetric semimonoidal comonads of Uustalu
and Vene \cite{comonads-notions}. The flat coeffect algebra is defined as $(\{1\}, \ast, \ast, \ast, 1, 1, =)$
where $1\ast1=1$ and $1=1$. The additional operation $\ident{merge}_{1,1}$ is provided by the
monoidal operation called \ident{m} by Uustalu and Vene. The $\ident{split}_{1,1}$ operation
is defined by duplication.
\end{example}

\begin{example}[Indexed Maybe comonad]
\label{thm:semantics-indexed-opt}

The flat coeffect algebra for liveness defines $\cpar$ and $\czip$, respectively as $\sqcup$ and $\sqcap$
and specifies that $\ident{D} \sqsubseteq \ident{L}$. Recall also that the object mapping is defined
as $\ctyp{\ident{L}}{\alpha} = \alpha$ and $\ctyp{\ident{D}}{\alpha} = 1$. The additional operations
of a flat indexed comonad are defined as follows:
%
\begin{equation*}
\begin{array}{rcl}
\ident{merge}_{\ident{L}, \ident{L}}~(a, b) &\narrow{=}& (a, b)\\
\ident{merge}_{\ident{L}, \ident{D}}~(a, ()) &\narrow{=}& ()\\
\ident{merge}_{\ident{D}, \ident{L}}~((), b) &\narrow{=}& ()\\
\ident{merge}_{\ident{D}, \ident{D}}~((), ()) &\narrow{=}& ()\\
\end{array}
\qquad
\begin{array}{rcl}
\ident{split}_{\ident{L}, \ident{L}}~(a, b) &\narrow{=}& (a, b)\\
\ident{split}_{\ident{L}, \ident{D}}~(a, b) &\narrow{=}& (a, ())\\
\ident{split}_{\ident{D}, \ident{L}}~(a, b) &\narrow{=}& ((), b)\\
\ident{split}_{\ident{D}, \ident{D}}~() &\narrow{=}& ((), ()))\\
\end{array}
\end{equation*}
\end{example}

@@

For implicit parameters, both $\czip$ and $\cpar$ are the $\cup$ operation and the relation
$\cleq$ is formed by the subset relation $\subseteq$. Recall that the comonadic data type $\ctyp{\cclrd{r}}{\alpha}$
is $\alpha \times (\cclrd{r} \rightarrow \ident{num})$ where $\ident{num}$ is the type of implicit parameter values.
The additional operations are defined as:
%
\begin{equation*}
\begin{array}{rcl}
\ident{split}_{\cclrd{r}, \cclrd{s}}~((a,b), g) &\narrow{=}& ((a, \restr{g}{\cclrd{r}}), (b, \restr{g}{\cclrd{s}}))\\
\ident{merge}_{\cclrd{r}, \cclrd{s}}~((a, f), (b, g)) &\narrow{=}& ((a, b), f \uplus g)\\
\end{array}
\quad
\begin{array}{l}
\textnormal{where}~f \uplus g = \\[-0.25em]
\quad\restr{f}{\, \textit{dom}(f) \setminus \textit{dom}(g)} \cup g
\end{array}
\end{equation*}
\end{example}

\noindent
The \ident{split} operation splits the tuple and restricts the function (representing available
implicit parameters) to the required subsets. The \ident{merge} operation is more
interesting. It uses the $\uplus$ operation that we defined when introducing implicit parameters
in Section~\ref{sec:applications-flat-impl}. It merges the values, preferring the definitions from
the right-hand side (call site) over left-hand side (declaration site). Thus the operation is not
symmetric.

@@

The $\ident{cobind}_{m,n}$ operation requires $m + n$ elements in order to generate $n$ past results
of the $f$ function, which itself requires $m$ past values. When combining two lists,
$\ident{merge}_{m,n}$ behaves as \emph{zip} and produces a list that has the length of the shorter
argument. When splitting a list, $\ident{split}_{m, n}$ needs the maximum of the required lengths.

@@

\begin{figure*}[t]
The semantics is defined over a typing derivation:
%
\begin{equation*}
\hspace{-3em}
\begin{array}{ll}
\hspace{5.2em}\semdef
  {\coctx{\Gamma}{\cunit} \vdash x_i : \tau_i }
  {\pi_i \circ \ident{counit}_{\cunit}}
& (\emph{var})
\\[1em]
\hspace{4.9em}\semdef
  {\coctx{\Gamma}{\czero} \vdash n : \ident{num} }
  {\ident{const}~n}
& (\emph{num})
\\[1.5em]
\hspace{2.2em}\semdeff
  {\coctx{\Gamma, x:\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
  {f}
  {f \circ \ident{curry}~\ident{merge}_{\cclrd{r}, \cclrd{s}} }
& (\emph{abs})
\\[2em]
\emdefff
  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2}
  {\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
  {\begin{array}{l}\sem{\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2}\\[-0.3em]~\end{array}}
  {f}
  {g}
  {\hspace{-0.5em}\begin{array}{l}
  \ident{app}~\circ~f\hspace{-0.15em}\times\hspace{-0.15em}(\ident{cobind}_{\cclrd{s}, \cclrd{t}}~g)~\circ~\ident{split}_{\cclrd{r}, \cclrd{s} \,\cseq\, \cclrd{t}} \\[-0.3em]
  ~~\circ~\ident{map}_{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})}~\ident{dup}\\[-0.3em]
  \end{array}\hspace{-0.5em}}
& (\emph{app})
\end{array}
\end{equation*}

@@

In Section~\ref{sec:applications-flat}, we defined the semantics of concrete (flat) context-dependent
computations including implicit parameters, liveness and data-flow. Using the \emph{flat indexed
comonad} structure, we can now define a single uniform semantics that is capable of capturing all
our examples, as well as various other computations.

As discussed in Section~\ref{sec:flat-unique}, different typing derivations of coeffect programs
may have different meaning (\eg~when working with implicit parameters) and so the semantics is
defined over a \emph{typing derivation} rather than over an \emph{term}. To assign a semantics to
a term, we need to choose a particular typing derivation. The algorithm for choosing a unique
typing derivation for our three systems has been defined in Section~\ref{sec:flat-unique}.

\paragraph{Contexts and types.}
The modelling of contexts and functions generalizes the concrete examples discussed in
Chapter~\ref{ch:applications}. We use the family of mappings $\ctyp{\cclrd{r}}{}$ as an (indexed)
data-type that wraps the product of free variables of the context and the arguments of functions:
%
\begin{equation*}
\begin{array}{rcl}
\sem{\coctx{x_1\!:\!\tau_1, \ldots, x_n\!:\!\tau_n}{ \cclrd{r} } \vdash e : \tau}
  &:& \ctyp{\cclrd{r}}{(\tau_1 \times \ldots \times \tau_n)} \rightarrow \tau\\
\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2} &=& \ctyp{\cclrd{r}}{\tau_1} \rightarrow \tau_2
\end{array}
\end{equation*}

@@

The semantics of variable access and abstraction are the same as in the semantics of Uustalu and
Vene \cite{comonads-notions}, modulo the indexing. The semantics of variable access (\emph{var}) uses
$\ident{counit}_{\cunit}$ to extract a product of free variables, followed by projection
$\pi_i$ to obtain the variable value. Abstraction (\emph{abs}) is interpreted as a curried
function that takes the declaration-site context and a function argument, merges them using
$\ident{merge}_{\cclrd{r}, \cclrd{s}}$ and passes the result to the semantics of the body $f$.
Assuming the context $\Gamma$ contains variables of types $\sigma_1, \ldots, \sigma_n$, this gives
us a value $\ctyp{\cclrd{r}\czip\cclrd{s}}((\sigma_1 \times \ldots \times \sigma_n) \times \tau_1)$.
Assuming that $n$-element tuples are associated to the left, the wrapped context is equivalent to
$\sigma_1 \times \ldots \times \sigma_n \times \tau_1$, which can then be passed to the body of the
function.

@@

\paragraph{Properties.}
The categorical semantics  in Section~\ref{sec:semantics-translation} defines a translation
that embeds context-dependent computations in a functional programming language, similarly to how
monads and the do notation provide a way of embedding effectful computations in Haskell.

@@

Informally, the following states that if we see the semantics as a translation, the resulting code
is well-typed. We revisit the property in Lemma~\ref{thm:semantics-welltyped} once we define
the target language and its typing.

\begin{lemma}[Correspondence]
\label{thm:flat-correspondence}
In the semantics defined in Figure~\ref{fig:semantics-flat}, the context annotations $\cclrd{r}$ of
typing judgements $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ and function types
$\tau_1 \xrightarrow{\cclrd{r}} \tau_2$  on the left-hand side correspond to the indices of mappings
$\ctyp{\cclrd{r}}{}$ in the corresponding semantic function on the right-hand side.
\end{lemma}
\begin{proof}
By analysis of the semantic rules in Figure~\ref{fig:semantics-flat}. We need to check that the
domains and codomains of the morphisms in the semantics (right-hand side) match.
\end{proof}

@@

\caption{Common syntax and reduction rules of the target language}
\label{fig:semantics-target}
\end{figure}

@@

\section{Translational semantics}
\label{sec:semantics-translation}

@@

\subsection{Functional target language}
\label{sec:semantics-translation-target}
